---
title: "seth_dice_4.2"
author: "Seth Dice"
date: "2025-06-29"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
# Assignment: 4.2 Assignment
# Name: Dice, Seth
# Date: 2025-06-29

suppressMessages(library(ggplot2))
suppressMessages(library(openxlsx))
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(cluster))
suppressMessages(library(factoextra))
suppressMessages(library(ggfortify))

setwd(
  paste0(
  "C:\\Users\\sethd\\OneDrive\\Desktop\\MS in Data Science Bellevue",
  "\\Classes\\DSC 630-T301 Predictive Analytics\\Week 4"
  ))

# Import ALS data
als_data <- read.csv("als_data.csv")

```

# Remove any data that is not relevant to the patient’s ALS condition.

Displaying column names from dataset to review which columns are not relevant to
a patient's ALS condition.

```{r, echo = FALSE}

# Show column names from dataset to determine which columns to remove
colnames(als_data)

## Using google to determine if the column is important for ALS
## Columns deemed not important - ID, Age_mean, Gender_mean, SubjectID.
## Age and gender are not important factors in ALS condition and ID/SubjectID
## are IDs given to the patients and not relevant to their ALS condition

als_data <- als_data %>%
  select(
    -ID, -Age_mean, -Gender_mean, -SubjectID
  )

```

The columns that are deemed not important to ALS condition per Google resources:

* ID - Column ID given to patient
* SubjectID - Column ID given to subject
* Age_mean - Age is not relevant factor in ALS condition
* Gender_mean - Gender is not relevant factor in ALS condition

# Apply a standard scalar to the data.

```{r, echo = FALSE}

# Using scale() to apply standard scalar to data
als_data_scalar <- scale(als_data)

```

Used R-function scale() to apply standard scalar to ALS data.

# Create a plot of the cluster silhouette score versus the number of clusters in a K-means cluster.

```{r, echo = FALSE}

# Creating function to compute average silhouette for k clusters
avg_sil <- function(k) {
  km <- kmeans(als_data_scalar, centers = k, nstart = 25) # K-means clustering
  ss <- silhouette(km$cluster, dist(als_data_scalar)) # Calculating silhouette scores
  mean(ss[, 3]) # Returning average silhouette scores
}

# Compute and plot k-means 2 through 15
k_val <- 2:15

# Extract average silhouette for 2-15 clusters
avg_sil_values <- suppressWarnings(map_dbl(k_val, avg_sil))

# Using function fviz_nbclust to plot optimum number of clusters
fviz_nbclust(als_data_scalar, kmeans, method = "silhouette")

```

# Use the plot created in (3) to choose an optimal number of clusters for K-means. Justify your choice.

The Average Silhouette Method determines how well each value lies within its cluster.
High average silhouette width indicates a good clustering, so from plot created in
(3) we'd want the number of clusters with the highest average silhouette width.
The highest average silhouette width is where number of clusters = 2.

# Fit a K-means model to the data with the optimal number of clusters chosen in part (4).

```{r, echo = FALSE}

## Fitting K-means model to data using 2 clusters
k2 <- kmeans(als_data_scalar, centers = 2, nstart = 25)

## Using fviz_cluster to visualize K-means model
fviz_cluster(k2, data = als_data_scalar)

```

# Fit a PCA transformation with two features to the scaled data.

```{r, echo = FALSE}

# Using leg/hands_range as my two features
# Fitting PCA transformation to two features and scaling data in process
pca_als <- prcomp(als_data[, c("leg_range", "hands_range")], center = TRUE, scale = TRUE)

```

The two features used from ALS data are: leg_range, hands_range

# Make a scatterplot of the PCA transformed data coloring each point by its cluster value.

```{r, echo = FALSE}

# Using 2 clusters as this value has been confirmed earlier in script
num_clusters <- 2

# Perform kmeans() on the pca_als to highlight clusters in graph
cluster_res <- kmeans(pca_als$x[, 1:2], centers = num_clusters)

# Adding the cluster_res to ALS dataframe as factor
als_data$cluster <- as.factor(cluster_res$cluster)

# Plot the PCA results, coloring by cluster
autoplot(pca_als, data = als_data, colour = 'cluster', shape = 'cluster', frame = TRUE)

```

# Summarize your results and make a conclusion.

After importing the ALS data and reviewing the columns the features removed were
ID, SubjectID, Age_mean, and Gender_mean as Google research was not showing these
as relevant to a patient's ALS condition. Once columns removed and scalar applied
the optimal number of clusters selected was 2 when using the average silhouette method.
A K-means model was then fitted with k = 2, and a PCA transformation was applied using
two features chosen at random: leg_range and hand_range which is shown as a scatterplot.
The K-means clustering graph shows the first cluster having a wider range of values
when compared to the second cluster. The PCA scatterplot has vectors pointing in the 
same direction which would give a positive correlation to leg_range and hands_range.
