---
title: "seth_dice_8.2"
author: "Seth Dice"
date: "2025-07-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}

options(scipen = 999)
options(warn = -1)

# Assignment: 8.2 Assignment
# Name: Dice, Seth
# Date: 2025-07-27

suppressMessages(library(ggplot2))
suppressMessages(library(openxlsx))
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(forecast))

# Data Cleaning -----------------------------------------------------------
setwd(paste0(
  "C:\\Users\\sethd\\OneDrive\\Desktop\\MS in Data Science Bellevue",
  "\\Classes\\DSC 630-T301 Predictive Analytics\\Week 8")
)

# Read in data of us retail sales
# total monthly retail sales in the US from January 1992 until June 2021
df_sales <- read.csv("us_retail_sales.csv")

```

# Plot the data with proper labeling and make some observations on the graph.

```{r, echo = FALSE}

# Pivoting data to make it easier to plot
df_pivot <- pivot_longer(
  df_sales,
  cols = c(2:13)
) %>%
  rename("MONTH" = name, "SALES" = value)

# Basic time series plot / view just numbers
ggplot(df_pivot, aes(x = YEAR, y = SALES)) +
  geom_line() +
  labs(x = "Years (1992 - 2021)", y = "Sales ($)") +
  theme_minimal()

# Creating custom order for months so legend is in correct order
df_pivot$MONTH <- factor(df_pivot$MONTH, levels = c(
  "JAN", "FEB", "MAR", "APR", "MAY", "JUN",
  "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"
))

## Plotting graph showing sales over time / factor of months
ggplot(data = df_pivot, aes(x = YEAR, y = SALES)) +
  geom_line(aes(colour = MONTH)) +
  scale_x_continuous(breaks = seq(1992, 2021, by = 1)) +
  labs(x = "Years (1992 - 2021)", y = "Sales ($)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

Looking at the time series graphs plotted we can see dips for the years 2007-2008
and 2020. The 2007-2008 dip can be explained by the market crash in '08. To explain
the drop in 2020 we have to view the data split by months and we can see the 2020
dip started in April which would mark the closure of a lot of businesses in the US
due to COVID.

# Split this data into a training and test set. Use the last year of data (July 2020 – June 2021) of data as your test set and the rest as your training set.

```{r, echo = FALSE}

# Training set
df_training <- df_pivot %>%
  filter(
    YEAR == "2020" & MONTH %in% c("JAN", "FEB", "MAR", "APR", "MAY", "JUN") |
      YEAR < "2020"
  )

# create time series training data
training_ts <- ts(df_training$SALES, frequency = 12, start = c(1992, 1))

# Test set
df_test <- df_pivot %>%
  filter(
    YEAR == "2021" & !is.na(SALES) |
      YEAR == "2020" & MONTH %in% c("JUL", "AUG", "SEP", "OCT", "NOV", "DEC")
      )

# create time series training data
test_ts <- ts(df_test$SALES, frequency = 12, start = c(2020, 7))

```

# Use the training set to build a predictive model for the monthly retail sales.

```{r, echo = FALSE}

# Using auto.arima() to have R pick correct specification
auto_arim_model = auto.arima(training_ts)

# Showing output of specifications
auto_arim_model

```

# Use the model to predict the monthly retail sales on the last year of data.

```{r, echo = FALSE}

# Predicting monthly sales for last year of data
forecast_values <- forecast(auto_arim_model, h = length(test_ts))

forecast_values

```

# Report the RMSE of the model predictions on the test set.

```{r, echo = FALSE}

model_rmse <- round(sqrt(mean((forecast_values$mean - test_ts) ^ 2)))
paste0("The RMSE of model predictions on test set: ", model_rmse)

```








