---
title: "seth_dice_milestone_5"
author: "Seth Dice"
date: "2025-08-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, echo = FALSE}

options(warn = -1)

# Assignment: Milestone 5
# Name: Dice, Seth
# Date: 2025-08-10

suppressMessages(library(ggplot2))
suppressMessages(library(openxlsx))
suppressMessages(library(tidyverse))
suppressMessages(library(lubridate))
suppressMessages(library(recommenderlab))
suppressMessages(library(naniar))
suppressMessages(library(nnet))
suppressMessages(library(caret))

# Set working directory of data
setwd("C:\\Users\\sethd\\OneDrive\\Desktop\\MS in Data Science Bellevue\\Classes\\DSC 630-T301 Predictive Analytics\\Milestone\\5")

```

# Data Import

Importing Diabetes data from Excel spreadsheet and reviewing data with head() confirming data has been imported before creating target and feature dataframes.

```{r, echo = FALSE}

# Import diabetes health data
dia_file <- read.csv("diabetes_012_health_indicators_BRFSS2015.csv")

# Change into singular dataframe for use
df <- dia_file %>%
  mutate(Diabetes_012 = as.factor(Diabetes_012))

head(df)

df_target <- df %>% select(Diabetes_012)

df_features <- df %>% select(-Diabetes_012)

```

# Data Review

Confirming that data is candidate for Multinomial Logistic Regression model

```{r, echo = FALSE}

# Confirming Diabetes_012 has three or more categories
paste0("The factor summary of Diabetes_012 is:") 
summary.factor(df_target)

```

The Diabetes_012 represents the target variable and is made up of three different categories

```{r, echo = FALSE}

# Listing data types of data without target variable
paste0("The structure of feature variables are:") 
str(df_features)

```

The feature variables are integers and considered continuous/categorical datatypes

Summary factor information was taken for all the feature variables as well but too large to post.

# Data Visuals

```{r, echo = FALSE}
# Quick way to visually review dataframe to ensure no rows have missing values
# Using gg_miss_var function to review missing values of dataframe while
# showing results

gg_miss_var(df_features, show_pct = TRUE) +
  labs(x = "Feature Variable", y = "Missing Proportion")
```

Confirming that there are no missing feature variable values by checking missing proportion which is showing 0.00 for all.

```{r, echo = FALSE}

# Show imbalance of target variable via bar chart
# Renaming target variables to values being measured
# Adding count of values
df_target_names <- df_target %>%
  group_by(Diabetes_012) %>%
  mutate(
    Diabetes_012 = case_when(
      Diabetes_012 == 0 ~ "No Diabetes/Only Pregnancy",
      Diabetes_012 == 1 ~ "Prediabetes",
      TRUE ~ "Has Diabetes"
    ),
    num_cnt = n()
  ) %>%
  ungroup() %>%
  unique() %>%
  rename("Participant Result" = Diabetes_012)

df_target_names

```

Confirming dataset is imbalanced with target variable being dominated by "No Diabetes/Only Pregnancy"

```{r, echo = FALSE}

# Plotting bar chart of data imbalance
ggplot(df_target_names, aes(x = reorder(`Participant Result`, -num_cnt),
                            y = num_cnt, fill = `Participant Result`)) +
  geom_bar(stat = "identity") +
  xlab("Participant Result") +
  ylab("Number of Participants") +
  theme(axis.text.x = element_text(size = 6))

```

Bar chart visually displaying the data imbalance of target variable.

# Model Building

Splitting data into training/test and displaying head().

```{r, echo = FALSE}

###### Creating data partition to separate data into train/test
df_index <- createDataPartition(df$Diabetes_012, p = .80, list = FALSE)
train <- df[df_index, ]
test <- df[-df_index, ]

paste0("Training dataset:")
head(train)

paste0("Testing dataset:")
head(test)

###### Setting the reference level to 0 for no diabetes
train$Diabetes_012 <- relevel(train$Diabetes_012, ref = "0")

```

Training multinomial model and displaying summary.

```{r, echo = FALSE}

###### Training the multinomial model
multinom_model <- multinom(Diabetes_012 ~ ., data = df)
###### Checking the model
summary(multinom_model)

####### Converting coefficients to odds by taking exponential of coefficients
odds_ratio = exp(coef(multinom_model))
odds_ratio

```

Coefficients have been converted to odds by taking exponential of the coefficients.

# Model Validation

```{r, echo = FALSE}

# Creating function to review p-values with TRUE showing p-values > 0.05
# and FALSE for p-values < 0.05
p_value_t_f <- function(x){
  zWald_modelo <- (summary(x)$coefficients / 
                                summary(x)$standard.errors)
  a <- t(apply(zWald_modelo, 1, function(x) {x < qnorm(0.025, lower.tail = FALSE)} ))
  b <- t(apply(zWald_modelo, 1, function(x) {x > -qnorm(0.025, lower.tail = FALSE)} ))
  ifelse(a == TRUE & b == TRUE, TRUE, FALSE)
}

# Predicting the values for train dataset
train$ClassPredicted <- predict(multinom_model, newdata = train, "class")
# Building classification table
tab <- table(train$Class, train$ClassPredicted)
# Calculating accuracy - sum of diagonal elements divided by total obs
tab_round <- round((sum(diag(tab))/sum(tab))*100,2)

```

The p-value confirmation where TRUE are for p-values > 0.05 and FALSE for p-values < 0.05

```{r, echo = FALSE}

p_value_t_f(multinom_model)

```

The target variables predicted are:

```{r, echo = FALSE}

tab

```

Confirming that target variables in dataset were all predicted:

```{r, echo = FALSE}

paste0("The percentage of target variables predicted for training dataset are: ", tab_round)

### Predicting the class on test data
# Predicting the class for test dataset
test$ClassPredicted <- predict(multinom_model, newdata = test, "class")
# Building classification table
tab <- table(test$Class, test$ClassPredicted)
# Calculating accuracy - sum of diagonal elements divided by total obs
tab_round <- round((sum(diag(tab))/sum(tab))*100,2)

```

Running the same predictions on test dataset we can see all the target variables have been predicted as well:

```{r, echo = FALSE}

tab

paste0("The percentage of target variables predicted for test dataset are: ", tab_round)

```
















